Turbo fast no compromise gravitational wave parameter estimation

### 2022-12-28 src/tex/ms.tex
Add algorithm
Stylize flow MC
Finish introduction

% Define document class
\documentclass[twocolumn]{aastex631}
\usepackage{showyourwork}
\usepackage[ruled,vlined]{algorithm2e}
\usepackage{color}

@@

@@

% Abstract with filler text
\begin{abstract}
We present a light-weighted, flexible, and high-performance framework to estimate
gravitational wave event parameters. By combining heterodyned likelihood,
automatically differentiable and accelerators compatible waveforms, normalizing
flow enhanced gradient-based MCMC, we achieve parameter estimation for real
events like GW150914 and GW170817 within a minute of sampling time with little to
no compromise. Our framework does not require pre-training or explicit
reparameterization and can be generalized to handle higher dimensional
problems. We also discuss as near real-time parameter estimation has been shown
to be possible by multiple groups, what are the trade-offs and future
developments the community should be aware of. Our code for generating the
manuscript and running the analysis is publically available on GitHub.
\end{abstract}

@@

% Brief description of PE and old generation of code
Parameter estimation (PE) is one of the most commonly performed data analysis
tasks in gravitational wave (GW) data analysis. The central question of PE is to
infer the parameters of the GW source given the strain data. The PE results in
the most recent catalog of GW events are powered by several community-developed
PE codes, including \texttt{lalinference}, \texttt{pycbc}, and \texttt{bilby}
\cite{Romero-Shaw:2020owr} \kw{cite}. These packages have been tested by a
number of groups and are well regarded as the standard tools.

% Emerging tools from different group
While the standard tools have passed many robustness tests, they are known to be
computationally intensive. The exact amount of time needed to analyze one event
depends on the length of the event. These community tools take about a week to
analyze shorter events such as binary black hole (BBH) mergers, whereas they can
take up to a month on a computer cluster to analyze longer events such as binary
neutron star (BNS) mergers. There are efforts from multiple groups developing
tools to speed up the PE process \kw{Cite}.

@@

In this work, we present a light-weighted, flexible, and high performance
framework to estimate GW event parameters. Our framework implements the
following major features to achieve its performance:
\begin{enumerate}
\item Differential waveform models
\item Normalizing flow enhanced MCMC sampler
\item Heterodyned likelihood
\item Native support for accelerator
\item JIT compilation
\end{enumerate}
\kw{Dingo can't do BNS yet}

@@

\begin{algorithm}
\caption{An algorithm with caption}\label{alg:cap}
\KwData{Initial position}
\KwResult{Chains, $\log{\mathcal{L}}$}
\emph{Training}\;
% \While{not converged}{
%     \For{n_loc}{

%     }
%     Local Sampling
% }
\emph{Production}\;
\end{algorithm}

@@

>>>>

% Define document class
\documentclass[twocolumn]{aastex631}
\usepackage{showyourwork}
\usepackage[ruled,vlined,linesnumbered]{algorithm2e}
\usepackage{algorithmic}
\usepackage{color}

@@

\SetKwInput{Parameters}{Parameters}
\SetKwInput{Variables}{Variables}

@@

% Abstract with filler text
\begin{abstract}
We present a light-weighted, flexible, and high-performance framework to
estimate gravitational wave event parameters. By combining heterodyned
likelihood, automatically differentiable and accelerators compatible waveforms,
normalizing flow enhanced gradient-based Markov Chain Monte Carlo (MCMC), we
achieve parameter estimation for real events like GW150914 and GW170817 within a
minute of sampling time. Our framework does not require pre-training or explicit
reparameterization and can be generalized to handle higher dimensional problems.
We also discuss as near real-time parameter estimation has been shown to be
possible by multiple groups, what are the trade-offs and future developments the
community should be aware of. Our code for generating the manuscript and running
the analysis is publically available on GitHub.
\end{abstract}

@@

% Brief description of PE
Parameter estimation (PE) is one of the most commonly performed data analysis
tasks in gravitational wave (GW) data analysis. The central question of PE is to
infer the parameters of a particular GW model given the strain data. In the
standard compact binary coalescence scenario, this could mean inferring
intrinsic parameters such the masses and spins of the event, as well as
extrinsic parameters such as sky localization and distance from Earth to the
event. PE is also used in testing general relativity (GR), which the main
question is then to check for deviations in parameters that are related to
non-GR modifications \kw{cite}. PE is a crucial step in GW science, since it translate
characteristics of the strain data into astrophysically relevant quantities that
can be used to constrain astrophysical phenomenon such as theories of binary
evolution \kw{cite}.

% Old codes have trouble handling next gen data
The PE results in the most recent catalog of GW events are powered by several
community-developed PE codes, including \texttt{lalinference}, \texttt{pycbc},
and \texttt{bilby} \cite{Romero-Shaw:2020owr} \kw{cite}. These packages have
been tested by a number of groups and are well regarded as the standard tools.
While the standard tools have passed many robustness tests, they are known to be
computationally intensive. The exact amount of time needed to analyze one event
depends on the length of the event. These community tools take about a week to
analyze shorter events such as binary black hole (BBH) mergers, whereas they can
take up to a month on a computer cluster to analyze longer events such as binary
neutron star (BNS) mergers. In the coming decade, there are planned upgrade for
existing facilities and next-generation detector such as the Einstein Telescope
(ET) \kw{cite} and the Cosmic Explorer (CE) \kw{cite}. These upgrades will
increase the sensitivity of the detectors and allow for the detection of more
events with better signal-to-noise ratio (SNR). The number of events that will
be detected in the coming decade is expected to grow from around a thousand per
year to over a million per year. This will put a significant strain on the
current PE tools, which are not designed to handle this many events.

% Effort going to new tools development.
In order to handle the future catalog, there are efforts from multiple groups
developing tools to speed up the PE process. This includes methods that employ
modern tools such as deep learning-based method pre-trained on a large
collection of waveforms \kw{cite}, as well as methods that simplify the
computational challenge by leveraging our knowledge about GW signals \kw{cite}.
While these methods are promising avenues to tackle standard GW analysis tasks
in the coming decade, particularly about events involving compact binary coalescence
(CBC) in GR, they capitalize assumptions that may not be valid for analysis involving
outliers and beyond-GR analysis.

@@

In this work, we present a light-weighted, flexible, and high performance
framework to estimate GW event parameters. Our framework implements the
following major features to achieve its performance:
\begin{enumerate}
\setlength{\itemsep}{0pt}
\item Differential waveform models
\item Normalizing flow enhanced MCMC sampler
\item Heterodyned likelihood
\item Native support for accelerator
\end{enumerate}
The main advantage of our framework is it does not rely on specific models or
assumptions about the problem to achieve its performance. This makes our method
extensible to problems beyond the standard CBC analysis.

@@

\begin{algorithm}
\caption{flowMC pseudocode}\label{alg:cap}
\KwIn{Initial position $ip$}
\Parameters{n training loop $nt$, n production loop $np$}
\Variables{current chain $cc$, current position $cp$, current NF parameters $\Theta$}
\KwResult{Chains}
$cp \leftarrow ip$\\
\tcc{Training loop}
    \For{$i<nt$}{
        $cc, cp \leftarrow LocalSampling(cp)$\\
        $\Theta \leftarrow TuneNF(cc)$
        $chains, cp \leftarrow GlobalSampling(cp, \Theta)$ 
        $cc \leftarrow Append(cc, chains)$
    }
\tcc{Production loop}
    \For{$i<np$}{
        $Chains, cp \leftarrow LocalSampling(cp)$
        $subchains, cp \leftarrow GlobalSampling(cp, \Theta)$ 
        $Chains \leftarrow Append(cc, chains)$
    }

\Return{$Chains$}
\end{algorithm}

@@

The two relevant works represent two orthogonal directions one can take in
building next generation tools in general. On one hand, there are modern tools
such as deep learning that are very flexible and powerful, but may need to rely
on having highly robust training data. On the other hand, there are traditional
tools that make use of our understanding of the underlying physics to simplify
the problem, which relies on having good intuition of the problem. Both
approaches rely on having a somewhat reasonable prior on how to approach the
problem. The main difference between methods used in industrial products and
scientific problems is scientific problems often try to address questions one
may have not been answered before, hence it is supposed to be far away from our
prior knowledge, meaning one need to take capability to generalize beyond the
current problem into account while building the tool. Our work utilizes both
reparameterization and deep learning, yet our method can be trivially extended
to problems beyond standard GW analysis that reparameterization or deep learning
alone may have trouble dealing with. We believe such flexibility is crucial for
building scientific tools.
